\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

>>>>

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- F
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

```{r  plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset. A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. D: The base series, a representative sample of the entire grey area, is represented by the green points."}

data("ERSPC")
#ogERSPC<-ERSPC
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))

toRemove<-sample(which(ERSPC$DeadOfPrCa==1),round(length(which(ERSPC$DeadOfPrCa==1))*0.7))
ERSPC$DeadOfPrCa[toRemove]<-0

a <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa")

plota<-emptyPlot<-plot(a,add.case.series=F,legend=F) +scale_y_continuous(labels = label_number_si())

plotb<-emptyPlot+geom_point(aes(x=a$time,y=a$ycoord, color = "Case series", fill = "Case series"), show.legend = F, size = 1.5, alpha = 0.5, shape = 21)+ theme(legend.position="none") +scale_y_continuous(labels = label_number_si())

plotc<-plot(a,add.case.series=T,legend=F) +scale_y_continuous(labels = label_number_si())

#TEMP MAINLY NEED FOR LEGEND
plotd<-plot(a,add.base.series=T) +scale_y_continuous(labels = label_number_si())

legend_b <- get_legend(
  plotd + 
    guides(color = guide_legend(nrow = 1)) +
     theme(legend.direction = "horizontal",legend.justification="center" ,legend.box.just = "bottom")
)

#remove legend
plotd<-plotd + theme(legend.position="none")
plotspop<-cowplot::plot_grid(plota,plotb,plotc,plotd,labels=c("A","B","C","D"))
g<-plot_grid(plotspop, legend_b, ncol = 1, rel_heights = c(1, .1))
g

```

@@

```{r plot-stratified-erspc-data, eval = F, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

Figure \ref{fig:plot-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis location of each case moment  is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area (Figure \ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-erspc-data} C, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time (Figure \ref{fig:plot-erspc-data} D). The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

\textcolor{red}{Poisson regression can be used to estimate the full hazard. However, this method requires the user to decide on the number of intervals. Too fine the intervals may result in many empty bins. This may cause the commonly used Newton-Raphson procedure to fail to converge [@kalbfleisch2011statistical]. If the intervals are too wide, the nonlinear trends that are present in the hazard may be masked. Rather than discretizing time like in Poisson regression, we sample person-specific moments which have been experienced during the study.}

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-stratified-erspc-data-1} \caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{Schunk}

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{stepwise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment in December 2003, the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

Next, layers for the case series and base series are added. The y-axis
location of each case moment is sampled at random vertically on the plot
to avoid having all points along the upper edge of the gray area. By
randomly distributing the cases, we can get a sense of the incidence
density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that
more events are observed at later follow-up times. Therefore, a constant
hazard model would not be appropriate in this instance as it would
overestimate the incidence earlier on in time, and underestimate it
later on. Finally, the base series is sampled horizontally with sampling
weight proportional to their follow-up time. The reader should refer to
the package vignettes for more examples and a detailed description of
how to modify the aesthetics of a population-time plot.

@@

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#>     data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.403  -0.172  -0.135  -0.085   3.801  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5787.6  on 54535  degrees of freedom
#> AIC: 5798
#> 
#> Number of Fisher Scoring iterations: 8
\end{Soutput}
\end{Schunk}

@@

absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\begin{Soutput}
#>  time                
#>     0 0.0e+00 0.0e+00
#>     1 5.4e-05 4.2e-05
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
#> NULL                               54539       6059             
#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
#> ScrArm              1        8     54535       5788   0.0048 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high        cov_prof
#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{table}

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
\end{figure}
\end{Schunk}

>>>>

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./plot-erspc-data-1} \caption[Population time plot for the ERSPC dataset]{Population time plot for the ERSPC dataset. A: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. B: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. C: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. D: The base series, a representative sample of the entire grey area, is represented by the green points.}\label{fig:plot-erspc-data}
\end{figure}
\end{Schunk}

@@

Figure \ref{fig:plot-erspc-data} is built sequentially by first adding a
layer for the area representing the population time in gray (Figure
\ref{fig:plot-erspc-data} A), with subjects having the least amount of
observation time plotted at the top of the y-axis. We immediately notice
a distinctive \emph{stepwise shape} in the population time area. This is
due to the randomization of the Finnish cohorts which were carried out
on January 1 of each of year from 1996 to 1999. Coupled with the uniform
December 31 2006 censoring date, this led to large numbers of men with
exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time
(i.e.~from right to left), the population-time plot shows the
recruitment pattern from its beginning in 1991, and the January 1
entries in successive years. Tracked forwards in time (i.e.~from left to
right), the plot for the first three years shows attrition due entirely
to death (mainly from other causes). Since the Swedish and Belgian
centres were the last to complete recruitment in December 2003, the
minimum potential follow-up is three years. Tracked further forwards in
time (i.e.~after year 3) the attrition is a combination of deaths and
staggered entries. As we can see, population-time plots summarise a
wealth of information about the study into a simple graph.

Next, layers for the case series and base series are added. The y-axis
location of each case moment is sampled at random vertically on the plot
to avoid having all points along the upper edge of the gray area (Figure
\ref{fig:plot-erspc-data} B). By randomly distributing the cases, we can
get a sense of the incidence density. In Figure
\ref{fig:plot-erspc-data} C, we see that more events are observed at
later follow-up times. Therefore, a constant hazard model would not be
appropriate in this instance as it would overestimate the incidence
earlier on in time, and underestimate it later on. Finally, the base
series is sampled horizontally with sampling weight proportional to
their follow-up time (Figure \ref{fig:plot-erspc-data} D). The reader
should refer to the package vignettes for more examples and a detailed
description of how to modify the aesthetics of a population-time plot.

@@

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 162 
#> Number of base moments: 16200 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#>     data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.268  -0.171  -0.133  -0.091   3.623  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)            -10.824      0.503  -21.52   <2e-16 ***
#> bs(Follow.Up.Time)1      2.639      1.359    1.94   0.0523 .  
#> bs(Follow.Up.Time)2      2.479      0.951    2.61   0.0092 ** 
#> bs(Follow.Up.Time)3      3.063      1.354    2.26   0.0237 *  
#> ScrArmScreening group   -0.207      0.161   -1.29   0.1986    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 1817.7  on 16361  degrees of freedom
#> Residual deviance: 1750.8  on 16357  degrees of freedom
#> AIC: 1761
#> 
#> Number of Fisher Scoring iterations: 8
\end{Soutput}
\end{Schunk}

@@

absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\begin{Soutput}
#>  time                
#>     0 0.0e+00 0.0e+00
#>     1 2.6e-05 2.1e-05
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
#> NULL                               16361       1818             
#> bs(Follow.Up.Time)  3     65.2     16358       1752  4.5e-14 ***
#> ScrArm              1      1.7     16357       1751      0.2    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high        cov_prof
#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
#> 2    1  2.6e-05  1.3e-05   5.7e-05   Control group
#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
#> 4    1  2.1e-05  9.6e-06   4.4e-05 Screening group
\end{Soutput}
\end{Schunk}

@@

@@
